Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-25 demystify.tex
various spelling error

\begin{enumerate}
  \item Should you apply?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty (\autoref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, statement of purpose (SOP), and  letters of recommendation (LORs).
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Pubication can greatly help}. Publications are good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-experiences}).

          \item \emph{Grades might be important}, depends on the reputation of \emph{your school} (\autoref{sec:your-school}). While good grades might not help much, bad ones will likley hurt your case.
          \item \emph{Standard tests are not important} (\autoref{sec:standard-tests}). GRE \emph{is not} even required. Spend your time on something else! For standard English tests (not required for domestic students), just do enough to pass the minimum requirements.

          \item \emph{Contacting a prof. for is recommended}, but do it \emph{properly} (\autoref{sec:contact}).

@@

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in the underepresentative group in CS can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipments and such for your research (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

\begin{commentbox}[Didier]
  \emph{Should letter writers have PhDs?}  In Rwanda, a lot of students interact more with teaching faculty who might not have PhD.
  \tcblower
  \textbf{Vu}: This is an interesting and useful detail that US faculty might not be aware of and students should mention about this in their statements. In general, someone with a PhD has been through the research process and therefore can better evaluate your research ability.  But if you do not have such people, then someone who can properly evaluate your research ability is still OK.
\end{commentbox}

@@

This is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part, so do not skip it.

@@

\paragraph{Follow Up Emails} If you had an interview and have not heard back, you can send an email to ask about the status of your application Don't send this too soon though, perhaps wait at least 2 weeks.

You can also send an email to ask about the status of your application if you have received offers from other schools and need to make a decision soon. See \autoref{sec:accept-postpone-decline} for how to accept/postpone/decline offers.

@@

In general, fellowships, especially those that are open to everyone, are highly competitive and prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only few  would have NSF Graduate or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even ``honorable mentioning'', school will still mention you on their website and you still should put it on your CV.

@@

While I find CSrankings useful, it is not perfect, and in general, rankings are superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should consider attending Open Houses (\autoref{sec:accepted}) and contact profs. that you're interested in at those place and talk to them. They would be more willing to chat with you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students. See more in \autoref{sec:accepted} on what to do after getting admitted.

@@

First, read the prof's website, see if they say something about contacting them. Many profs. explicitly indicate how prospective students should (or should not) contact them (e.g., using specific email subjects).
In general, the best way to catch the prof.'s attention is to \emph{customize your email} for them.  For example, read their papers, know what they work on, and see if you are interested in their research. Then send them an email talking how/why their work would match yours.
In contrast, if you write a generic email that can be sent to multiple professors (e.g., if you just change some names and keywords in the email or copy and paste paper titles), you will not get a response.

@@

\begin{commentbox}[Vu:]
  I would strongly advocate for a non-STEM student who shows that they have a strong drive for CS by studying core CS knowledge through various channels (e.g., self-study through online courses, projects, etc).  I have seen many students with non-CS background who are very successful in CS PhD.  I also have seen many students with CS background who are not successful in CS PhD.  So it is not about your background, it is about your drive and passion for CS research.
\end{commentbox}

@@

The first 2 years you typically take coursework (somewhat equivalent to an MS study), find an adviser, learn how to do research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job.
In many cases you might take a summer or two off to do internship to get additional research opportunities.
The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year where the ambition is \emph{``Just let me graduate''}).

@@

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$29000 and a 3-month summer of \$10000 (about a third of the 9-month stipend). I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., Univ. of Chicago is a whooping \$70K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

>>>>

\begin{enumerate}
  \item Should you apply?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members (\autoref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, statement of purpose (SOP), and  letters of recommendation (LORs).
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Publications can greatly help}. Publications are good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-experiences}).

          \item \emph{Grades might be important}, depends on the reputation of \emph{your school} (\autoref{sec:your-school}). While good grades might not help much, bad ones likely will hurt your case.
          \item \emph{Standard tests are not important} (\autoref{sec:standard-tests}). GRE \emph{is not} even required. Spend your time on something else! For standard English tests (not required for domestic students), just do enough to pass the minimum requirements.

          \item \emph{Contacting a prof. is recommended}, but do it \emph{properly} (\autoref{sec:contact}).

@@

    \item Miscs and FAQS
        \begin{itemize}
          \item Why you were rejected (\autoref{sec:why-rejected})
          \item Increasing admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item Being in the underepresentative groups in CS can help (\autoref{sec:urm}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item You do not need an MS degree (\autoref{sec:msrequirement}) to apply for PhD in CS, and it takes about 5--7 years for PhD in CS (\autoref{sec:time}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item How do you call a professor? (\autoref{sec:address})
          \item Your entire PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost})
          \item Ask your prof. if they can buy computer equipments and such for your research (\autoref{sec:buying-equipment}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

\begin{commentbox}[Didier]
  \emph{Should letter writers have PhDs?}  In Rwanda, a lot of students interact more with teaching faculty who might not have PhD.
  \tcblower
  \textbf{Vu}: This is an interesting and useful detail that US faculty might not be aware of and students should mention this in their statements. In general, someone with a PhD has been through the research process and therefore can better evaluate your research ability.  But if you do not have such people, then someone who can properly evaluate your research ability is still OK.
\end{commentbox}

@@

This is something easy to do, but is neglected by many
applicants: \textbf{customize the statement} for the school you're applying to,
e.g., why do you apply \emph{here}? provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part (\autoref{sec:why-rejected}), so do not skip it.

@@

\paragraph{Follow-Up Emails} If you had an interview and have not heard back, you can email to ask about the status of your application Don't send this too soon though, perhaps wait at least 2 weeks.

You can also email to ask about the status of your application if you have received offers from other schools and need to make a decision soon. See \autoref{sec:accept-postpone-decline} for how to accept/postpone/decline offers.

@@

In general, fellowships, especially those that are open to everyone, are highly competitive and prestigious, and you will stand out if you get one.  Every PhD student has pubs, but only few  would have NSF Graduate or Microsoft fellowship. In fact, these are so prestigious that even if you didn't get it but make it to the final round or even ``honorable mentioning'', school will still mention you on their website, and you should put it on your CV.

@@

While many find CSRankings useful, it is not perfect, and in general, rankings are superficial, and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should consider attending Open Houses (\autoref{sec:accepted}) and contact profs. that you're interested in at those places and talk to them. They would be more willing to chat with you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students. See more in \autoref{sec:accepted} on what to do after getting admitted.

@@

First, read the prof's website, see if they say something about contacting them. Many profs. explicitly indicate how prospective students should (or should not) contact them (e.g., using specific email subjects).
In general, the best way to catch the prof.'s attention is to \emph{customize your email} for them.  For example, read their papers, know what they work on, and see if you are interested in their research. Then email them and talk about how/why their work would match yours.
In contrast, if you write a generic email that can be sent to multiple professors (e.g., if you just change some names and keywords in the email or copy and paste paper titles), you will not get a response.

@@

\begin{commentbox}[Vu:]
  I would strongly advocate for a non-STEM student who shows that they have a strong drive for CS by studying core CS knowledge through various channels (e.g., self-study through online courses, projects, etc.).  I have seen many students with non-CS background who are very successful in CS PhD.  I also have seen many students with CS background who are not successful in CS PhD.  So it is not about your background, it is about your drive and passion for CS research.
\end{commentbox}

@@

The first 2 years you typically take coursework (somewhat equivalent to an MS study), find an adviser, learn how to do research.  The next 2--3 years you focus on your research, form dissertation topic, and get results published. The last 1-2 years you continue to publish, write and defend your dissertation, and look for job.
In many cases you might take a summer or two off to do internship to get additional research opportunities.
The PhDComics figure on top shows the ``ambition'' level of a PhD student over their years of study (they miss the 6--7th year when the ambition is \emph{``Just let me graduate''}).

@@

\autoref{tab:cost} shows the budget breakdown for a GRA per year.
These numbers are based on my experience at public universities in the US. Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$29000 and a 3-month summer of \$10000 (about a third of the 9-month stipend). I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., Univ. of Chicago is a whooping \$70K) and a 58.9\% rate on \emph{indirect cost}, which is a typical rate charged for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the students take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).
